Safe Transferable Regions

### 2015-11-17 paper/abstract.tex
Edits to Sec.1

Manual memory management is susceptible to the usual perils of dangling pointers.
A key contribution of this paper is a region type system that
ensures the memory safety of region-annotated C\# programs,
even in presence of higher-order stateful functions. We
also present a type inference algorithm that infers principal region types
for first-order programs, eliminating the need for programmers to write
region annotations on types. Our inference algorithm is also very
effective, albeit incomplete, for higher-order programs.

>>>>

Manual memory management is susceptible to the usual perils of
dangling pointers.  A key contribution of this paper is a region type
system that ensures the memory safety of C\# programs in the presence
of transferable regions. We compliment our type system with a type
inference algorithm that infers principal region types for first-order
programs. This eliminates the need for programmers to write region
annotations on types, while facilitating the reuse of existing C\#
libraries with no modifications. Various case studies demonstrate the
practical utility of our approach.

### 2015-11-17 paper/introduction.tex
Edits to Sec.1

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer.  In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, the garbage collection overhead becomes
significant.
% as the collector scans these large data-structures
However, these data-structures have a well-defined lifetime, which can
be exploited to achieve better performance.
% using alternative memory management techniques.

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow).
% that is used to realize a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is freed in one shot.
In this example, the output to be constructed for each time window $t$ can be
implemented using a separate region.

@@

The goal of this work is a \emph{memory-safe} region-based memory management technique
that supports dynamic regions (that are first-class objects) and higher-order functions.
Though we focus, in this paper, on memory-safety, our approach offers other benefits.
When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
provides an ownership discipline that ensures that the producer cannot subsequently modify
the data-structure. When the same transfer happens in a distributed setting, the system guarantees
that the data-structure is self-contained and can be safely transmitted across address spaces.
We refer to dynamic regions that provide these guarantees as \emph{transferable} regions.

@@

We use \emph{stack regions} (regions with lexically-scoped lifetimes), as well as the stack,
for such temporary pointers (to objects inside a dynamic region) and use the following protocol
to ensure safety of such pointers.
To work with a dynamic region $D$, we must first \emph{open} region $D$,
to indicate that $D$ should not be freed during this period.
When the processing of the region is complete, we close the region $D$,
to indicate that it is safe to free $D$.
In general, a dynamic region may be opened and closed multiple times before it is finally freed.

@@

Dynamic regions are first-class objects in our system: \eg, in our running example,
we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
aliases (pointers) to the same region, which, in turn, means that statically checking if the program correctly follows the
dynamic region protocol is hard (undecidable, in fact). Safety can be ensured by using, \eg, linear typing to prevent the
creation of aliases, but this would be quite restrictive. Hence, in our system, we dynamically check for adherence to
this protocol.

@@

\begin{itemize}
  \item We present \name, a \csharp-like typed object-oriented language
  that eschews garbage collection in favour of programmer-managed
  memory regions . \name extends its core language, which includes
  \emph{lambdas} (higher-order functions) and \emph{generics}
  (parametric polymorphism), with constructs to create, manage and
  destroy static and dynamic memory regions. Dynamic regions are
  first-class values in \name.
  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
  % memory regions outperform their GC counterparts by a margin of upto
  % 59\%.

  \item \name is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that the region usage correctly follows a simple discipline.
  The correctness of region usage is checked efficiently at run-time.
  The overhead of checking these conditions is less than
  \GK{x\%} in our experiments with \naiad workloads.

  \item We define an operational semantics for \name, and a type
  safety result that formalizes and proves safety guarantees described
  above.

  \item We describe a region type inference algorithm for \name that
  (a). completely eliminates the need to annotate \name programs with
  region types, and (b). enables seamless interoperability between
  region-aware \name programs and legacy standard library code that is
  region-oblivious. The cornerstone of our inference algorithm is
  \csolve, a novel constraint solving algorithm that performs
  abduction and deduction in a Herbrand constraint system to solve the
  constraints generated by the type inference algorithm.

  \item We prove that region type inference is sound with respect to
  its type system, and also complete for the set of first-order
  programs (i.e., programs without lambdas).

>>>>

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer. In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, garbage collection overhead becomes
significant. Importantly, much of this overhead results from the
collector performing avoidable or unproductive
work~\cite{Broom:HotOS}. For example, a consumer process may trigger a
GC as soon as it receives a large data-structure from a producer
process, which has just finished its GC. In such case, consumer's GC
traverses the entire data-structure performing little or no work.
Another example is the process executing the code in
Fig.~\ref{fig:motivating-eg}, whose GC might repeatedly traverse the
\C{map} data-structure, although its objects cannot be collected until
a suitable timing message arrives.
% as the collector scans these large data-structures For instance, in
% the above example, the list at \C{map[t]} cannot be collected until
% the the timing message arrives, and the list is transferred.
% \C{transfer} operation is complete.  using alternative memory
% management techniques.

An important observation, in the context of processes of kind
described above, is that the data-structures exchanged between them
have can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline. A region is a block of memory that is allocated
and freed in one shot, consuming constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a set of fate-sharing objects.
% that is used to realize a data-structure with a well-defined
% lifetime. 
In this example, the output to be constructed for each time window $t$
can be implemented using a separate region.

@@

The goal of this work is a \emph{memory-safe} region-based memory
management technique that supports dynamic regions as first-class
objects. Though we focus, in this paper, on memory-safety, our
approach offers other benefits.  When a producer transfers a
data-structure to a consumer, in a shared-memory context , our
approach provides an ownership discipline that ensures that the
producer cannot subsequently modify the data-structure. When the same
transfer happens in a distributed setting, the system guarantees that
the data-structure is self-contained and can be safely transmitted
across address spaces.  We refer to dynamic regions that provide these
guarantees as \emph{transferable} regions.

@@

We use \emph{stack regions} (regions with lexically-scoped lifetimes),
as well as the stack, for such temporary pointers (to objects inside a
dynamic region) and use the following protocol to ensure safety of
such pointers.  To work with a dynamic region $D$, we must first
\emph{open} region $D$, to indicate that $D$ should not be freed
during this period.  When the processing of the region is complete, we
\emph{close} the region $D$, to indicate that it is safe to free $D$.
In general, a dynamic region may be opened and closed multiple times
before it is finally freed.

@@

Dynamic regions are first-class objects in our system. For instance,
in our running example, we would like to use a dictionary that stores
dynamic regions. However, this means that the program may create
multiple aliases (pointers) to the same region, which, in turn, means
that statically checking if the program correctly follows the dynamic
region protocol is hard (undecidable, in fact). Safety can be ensured
by using, \eg, linear typing to prevent the creation of aliases, but
this would be quite restrictive. Hence, in our system, we dynamically
check for adherence to this protocol.

@@

\begin{itemize} \item We present \name, a \csharp-like typed
object-oriented language that eschews garbage collection in favour of
programmer-managed memory regions . \name extends its core language,
which includes \emph{lambdas} (higher-order functions) and
\emph{generics} (parametric polymorphism), with constructs to create,
manage and destroy static and dynamic memory regions. Dynamic regions
are first-class values in \name.
  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
  % program using programmer-managed memory regions outperform their
  % GC counterparts by a margin of upto 59\%.

  \item \name is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that certain typestate invariants on regions hold.  The
  later invariants are enforced via simple runtime checks.
% The correctness of region usage is checked efficiently at run-time.
% The overhead of checking these conditions is less than \GK{x\%} in
% our experiments with \naiad workloads.

  \item We define an operational semantics for \name, and a type
  safety result that clearly defines and proves safety guarantees
  described above.

  \item We describe a region type inference algorithm for \name that
  (a). completely eliminates the need to annotate \name programs with
  region types, and (b). enables seamless interoperability between
  region-aware \name programs and legacy standard library code that is
  region-oblivious. The cornerstone of our inference algorithm is a
  novel constraint solver that performs abduction in a partial-order
  constraint domain to infer weakest solutions to recursive
  constraints.

### 2015-11-17 paper/overview2.tex
updates to section 3
Edits to Sec.1

\section{Overview of \name}
\label{sec:overview}

@@

\paragraph{Stack Regions}
The ``\C{letregion R S}'' construct creates a new stack region,
with a static identifier \C{R}, whose scope is restricted to the
statement \C{S}. The semantics of  \C{letregion} is similar
to Tofte and Talpin~\cite{tofte94}'s \C{letregion} expression:
new objects can be allocated in the newly created region within
the scope of \C{S}, but the region and all objects allocated within it
are freed at the end of \C{S}.

\paragraph{Object Allocation}
The ``\C{new@R T()}'' construct creates a new object of type \C{T} in
the region \C{R}. The specification of the allocation region \C{R} in
this construct is optional.  At runtime, \name maintains a stack of \emph{active}
regions, and we refer to the region at the top of the stack as the \emph{allocation
context}. The statement \C{new T()} allocates the newly created object in
the current allocation context.
%
This is important as it enables \name applications to use existing region-oblivious
C\# libraries. In particular, given a  C\# library function \C{F} (that makes no use of
\name's region constructs), the statement ``\C{letregion R F()}'' invokes \C{F},
but has the effect that all objects allocated by this invocation are allocated in the
new region \C{R}.

>>>>

\section{An Informal Overview of \name}
\label{sec:overview}

@@

\paragraph{Stack Regions}
The ``\C{letregion R \{ S \}}'' construct creates a new stack region,
with a static identifier \C{R}, whose scope is restricted to the
statement \C{S}. The semantics of  \C{letregion} is similar
to Tofte and Talpin~\cite{tofte94}'s \C{letregion} expression:
new objects can be allocated in the newly created region within
the scope of \C{S}, but the region and all objects allocated within it
are freed at the end of \C{S}.

\paragraph{Object Allocation}
The ``\C{new@R T()}'' construct creates a new object of type \C{T} in
the region \C{R}. The specification of the allocation region \C{R} in
this construct is optional.  At runtime, \name maintains a stack of \emph{active}
regions, and we refer to the region at the top of the stack as the \emph{allocation
context}. The statement \C{new T()} allocates the newly created object in
the current allocation context.
%
This is important as it enables \name applications to use existing
region-oblivious C\# libraries. In particular, given a  C\# library
function \C{f} (that makes no use of \name's region constructs), the
statement ``\C{letregion R \{ f(); \}}'' invokes \C{f}, but has the
effect that all objects allocated by this invocation are allocated in
the new region \C{R}.
